\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 paper.qmd
Added sumatised abstract taking by earlier ones

---
title: "Visualising How Non-linear Dimension Reduction Warps Your Data"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional data. However, their complexity and parameter choices may lead to distrustful or misleading results. To address this challenge, we propose a novel approach that combines the tour technique with a low-dimensional manifold generated using NLDR techniques, hexagonal binning, and triangulation. This integration enables a clear examination of the low-dimensional representation in the original high-dimensional space. Our approach not only preserves the advantages of both tours and NLDR but also offers a more intuitive perception of complex data structures and facilitates accurate data transformation assessments. The method and example data sets are available in the **quollr** R package.

>>>>

---
title: "Visualising How Non-linear Dimension Reduction Warps Your Data"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

### 2024-03-26 paper.tex
Added sumatised abstract taking by earlier ones

\bigskip
\bigskip
\begin{abstract}
Non-Linear Dimension Reduction (NLDR) techniques have emerged as
powerful tools to visualize high-dimensional data. However, their
complexity and parameter choices may lead to distrustful or misleading
results. To address this challenge, we propose a novel approach that
combines the tour technique with a low-dimensional manifold generated
using NLDR techniques, hexagonal binning, and triangulation. This
integration enables a clear examination of the low-dimensional
representation in the original high-dimensional space. Our approach not
only preserves the advantages of both tours and NLDR but also offers a
more intuitive perception of complex data structures and facilitates
accurate data transformation assessments. The method and example data
sets are available in the \textbf{quollr} R package.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced]}{\end{tcolorbox}}\fi

>>>>

\bigskip
\bigskip
\begin{abstract}
Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional (high-D)
data using non-linear transformation. The methods and parameter choices
can create wildly different representations, making it difficult to
decide which is best, or whether any or all are accurate or misleading.
NLDR often exaggerates random patterns, sometimes due to the samples
observed. But NLDR views have an important role in data analysis
because, if done well, they provide a concise visual (and conceptual)
summary of high-D distributions. To help evaluate the NLDR we have
developed an algorithm to show the 2D NLDR model in the high-D space,
viewed with a tour. One can see if the model fits everywhere or better
in some subspaces, or completely mismatches the data. It is used to
evaluate which 2D layout is the best representation of the high-D
distribution and see how different methods may have similar summaries or
quirks.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
